---
title: "Data Validation"
subtitle: "_Heliconia_ Demographic Survey Data"
author: Eric R. Scott & Emilio M. Bruna
date: "updated: `r Sys.Date()`"
# output: html_document
  # pdf_document: default
  # html_document:
  #   df_print: paged
output: 
  rmdformats::robobook
#   # rmdformats::downcute:
# layout: page
# permalink: survey_validation/survey-validation
self_contained: true
thumbnails: true
lightbox: true
gallery: false
highlight: tango
bookdown: true
---

We use the R package [`pointblank`](https://rich-iannone.github.io/pointblank/) to review and validate the plot-level descriptors (`HDP_plots.csv`) and clean demographic data set (`heliconia_survey_clean.csv`) in preparation for archiving in Dryad and publication in Bruna et al. (2023). The report below includes:

1. the different validation tests that were conducted, 
1. the date of the most recent test, 
1. each test's criteria for 'pass', 'warn' and 'stop', 
1. the number of 'units' (i.e., rows or columns) assessed in each test, 
1. how many of these units passed or failed, and
1. a button for downloading a .csv file of the records flagged by a particular validation test. _Note that these are not necessarily errors_. For instance, the validation procedure for 'plant size - height' returns as 'stop' all plants >2 m tall. _Heliconia_ plants can exceed this threshold; this test is simply designed to flag any such individuals. In contrast, the data set should not have any duplicated rows. A notification of 'fail' for this test indicates an error that can be corrected by downloading the csv file, reviewing the duplicated rows, and uploading the necessary corrections. 

Last run: `r Sys.Date()`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

library(pointblank)
library(readr)
library(here)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(gghighlight)
# Use `validate_rmd()` here to set options for the
# pointblank validation workflow within R Markdown documents
```


```{r load-data}
ha <- read_csv(
  # here("data", "survey_archive", "HDP_1998_2009.csv")
  here("data", "survey_clean", "heliconia_survey_clean.csv")

)
ha_plots <- read_csv(
  here("data", "survey_archive", "HDP_plots.csv")
)
```


```{r al, echo=FALSE}
al_default <-  action_levels(warn_at = 1, stop_at = 0.02) #warn if even row fails, error if 2% of rows fail
al_strict <- action_levels(stop_at = 1) #error if even one row fails
```


```{r funs}
#calculate percent changes
ha_change <- 
  ha %>%
  group_by(plant_id) %>%
  arrange(plant_id, year) %>%
  mutate(ht_prev = lag(ht),
         shts_prev = lag(shts),
         ht_diff = ht - ht_prev,
         shts_diff = shts - shts_prev,
         ht_pc = abs(ht_diff / ht_prev),
         shts_pc = abs(shts_diff / shts_prev)) %>%
  #re-order columns so easier to see errors
  select(plot_id, subplot, plant_id, year, shts_prev, shts, shts_diff, shts_pc, ht_prev, ht, ht_diff, ht_pc, everything()) %>% 
  ungroup()
```

-----

# Dataset Structure: _Data types_

Tests to determine if columns are correctly coded as integer, character, etc.   
**Test criteria**: Strict ('stop' if *any* rows fail).

```{r agent1}
ha %>%
  create_agent(
    actions = al_default,
    label = "Data Validation"
  ) %>%
  col_vals_expr(label = "Height is measured to nearest cm",
                expr = ~ ht %% 1 == 0, actions = al_strict) %>%
  col_vals_expr(label = "Shoots is interger",
                expr = ~ shts %% 1 == 0, actions = al_strict) %>%
  col_vals_expr(label = "Number of inflorescences is integer",
                expr = ~ infl %% 1 == 0, actions = al_strict) %>% 
  interrogate()
```


-----

# Dataset Structure: _Plot & Subplot IDs_

Test for any nonexistent values of `plot_id` (e.g., 'FF-10', 'CF-23') or `subplot` (e.g., 'H23', 'A11').  
**Test criteria**: Strict ('stop' if *any* rows fail).


```{r agent2}
ha %>%
  create_agent(
    actions = al_default,
    label = "Data Validation"
  ) %>%
  col_vals_in_set(vars(plot_id), ha_plots$plot_id,
                  actions = al_strict) %>%
  col_vals_in_set(vars(subplot), unlist(lapply(LETTERS[1:10], paste0, 1:10)),
                  actions = al_strict) %>%
  interrogate()
```


# Dataset Structure: _Duplicated or Missing Values_

Tests for duplicated rows, missing `plant_ID` numbers, or duplicate `plant_id` numbers (test is done for every survey year).  
**Test criteria**: Strict ('stop' if *any* rows fail).

```{r agent3}
ha %>%
  create_agent(
    actions = al_default,
    label = "Data Validation"
  ) %>%
  rows_distinct(actions = al_strict, 
                label  = "duplicated rows") %>% 
  col_vals_not_null(vars(plant_id), actions = al_strict) %>% 
  rows_distinct(columns = "plant_id",
                preconditions = ~ . %>%
                  dplyr::group_by(plant_id,tag_number) %>%
                  slice(1),
                # segments = vars(year),
                actions = al_strict,
                label = "Check for duplicate plant ID numbers") %>% 
  rows_distinct(columns = "tag_number",
                preconditions = ~ . %>% 
                  dplyr::select(plot_id,plant_id,tag_number,subplot) %>%
                  unique() %>%
                  ungroup() %>%
                  group_by(plot_id,tag_number,.drop = FALSE) %>%
                  filter(n() > 1) %>% 
                  arrange(plot_id,tag_number),
                # segments = vars(plot_id),
                actions = al_strict,
                label = "Check for duplicate tag numbers in a plot") %>% 
  interrogate()
```


# Plant Characteristics: _Size & Flowering_

Tests to determine how many values of plant size (`shts`, `ht`) or infloresence number (`infl`) are outside the range of most values.  
**Test criteria**: 'warn' if $\geq$ 1 rows fail conditions, 'stop' if $\geq$ 2% of rows fail conditions.  

```{r agent4}
ha %>%
  create_agent(
    actions = al_default,
    label = "Data Validation"
  ) %>%
  # check values are within reasonable range
  col_vals_between(vars(shts), 0, 20, na_pass = TRUE,
                   label = "shoots between 0 and 20") %>%
  col_vals_between(vars(ht), 0, 200, na_pass = TRUE,
                   label = "height between 0 and 200cm") %>%
  col_vals_between(vars(infl), 0, 3, na_pass = TRUE,
                   label = "infloresences between 0 and 3") %>%
  interrogate()
```


# Plant Characteristics: _Growth_

Tests for unusual changes in plant size (both height and shoot number) from $Year_{t}$ to $Year_{t+1}$.  
**Test criteria**: 'warn' if $\geq$ 1 rows fail conditions, 'stop' if $\geq$ 2% of rows fail conditions.  


```{r agent5}
ha_change %>%
  create_agent(
    actions = al_default,
    label = "Check growth & regression"
  ) %>%
  
  # check that year-to-year change is reasonable
  col_vals_lt(vars(ht_pc), 2, na_pass = TRUE,
              label = "|% change in height| < 200%") %>%
  col_vals_between(vars(ht_diff), -100, 100, na_pass = TRUE,
                   label = "|∆ height| < 100cm",
                   # segments = vars(ht),
                   actions = al_strict) %>%
  col_vals_between(vars(shts_diff), -5, 5, na_pass = TRUE,
                   label = "|∆ shoot number| < 5",
                   # segments = vars(shts),
                   actions = al_strict) %>%
  interrogate()
```

# Seedlings: _Initial size_
Tests for seedlings whose size at initial marking was unusually large. Conducted for both height and shoot number.  
**Test criteria**: 'warn' if $\geq$ 1 rows fail conditions, 'stop' if $\geq$ 2% of rows fail conditions.  


```{r agent6}
# check seedlings
ha %>%
  filter(recorded_sdlg) %>% 
  create_agent(
    actions = al_default,
    label = "Check seedlings"
  ) %>% 
  col_vals_lt(vars(shts), 3, na_pass = TRUE,
              label = "shoots < 3") %>% 
  col_vals_lt(vars(ht), 30, na_pass = TRUE,
              label = "height < 30cm") %>% 
  interrogate()
```


# Seedlings: _Data Entry Errors_

Check if during data entry the size of seedlings (1) wasn't accidentally transposed to the "inflorescences" column, which would code a new seedling as being reproductive. 

**Test criteria**: Strict ('stop' if *any* rows fail).


```{r agent7}
# check seedlings
ha %>%
  filter(recorded_sdlg) %>% 
  create_agent(
    actions = al_strict,
    label = "Check for 'reproductive' seedlings"
  ) %>%
  col_vals_lt(columns = vars(infl), value = 1,
              na_pass = TRUE,
              label = "infl < 1",
              # segments = vars(shts),
              actions = al_strict) %>%
  interrogate()
```


------

<!-- # Missing values: _Height_ -->

<!-- Graphical depiction of the proportion of plants in each demographic plot for which there is no measurement of plant height (e.g., if plant not found). -->

<!-- ```{r echo=FALSE} -->
<!-- ha %>%  -->
<!--   group_by(plot_id, year) %>%  -->
<!--   summarize(prop_na = sum(is.na(ht))/n(), .groups = "drop") %>%  -->
<!--   ggplot(aes(x = year, y = prop_na)) + -->
<!--   geom_point(color="red") + -->
<!--   facet_wrap(~plot_id) + -->
<!--   theme_bw() + -->
<!--   scale_x_continuous(breaks = scales::breaks_pretty()) + -->
<!--   scale_y_continuous(breaks = c(0,0.5,1)) + -->
<!--   theme(axis.text.x = element_text(angle = 45, hjust = 1)) + -->
<!--   gghighlight(prop_na > .5) + -->
<!--   labs(y = "Proportion of values missing - Height") -->

<!-- ``` -->

# Zombie plants

Zombie plants are those that were recorded as 'Dead' in a survey but for which there is a measurement in a subsequent year (indicative of the plant losing all below-ground parts and then new shoots emerging prior to the next survey). This validation generates a `.csv` of any plants meeting this condition (labeled as 'zombie` for review and correction. 

```{r agent8}

zombies <- ha %>% 
  group_by(plant_id) %>%
  mutate(
    zombie = if_else(lag(census_status, 1) == "dead", "delete", NA_character_),
    .before = 1
    ) %>%
  fill(zombie, .direction = "down") %>%
  filter(zombie == "delete") %>%
  mutate(zombie = if_else((is.na(shts) & is.na(ht)), "delete", "zombie"))
zombies %>%
  create_agent(
    actions = al_default,
    label = "Check for zombies"
    ) %>%
  col_vals_equal(
    columns = vars(zombie),
    value = "zombie",
    na_pass = TRUE,
    actions = al_strict,
    label = "Check for Zombies"
    ) %>% 
  interrogate()
  
```


<!-- # Check for missing values: Shoots -->

```{r echo=FALSE}
# ha %>% 
#   group_by(plot_id, year) %>% 
#   summarize(prop_na = sum(is.na(shts))/n(), .groups = "drop") %>% 
#   ggplot(aes(x = year, y = prop_na)) +
#   geom_point(color="red") +
#   facet_wrap(~plot_id) +
#   theme_bw() +
#   scale_x_continuous(breaks = scales::breaks_pretty()) +
#   scale_y_continuous(breaks = c(0,0.5,1)) +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#   gghighlight(prop_na > .5) +
#   labs(y = "Proportion values missing - Shoots")
  
```



# Plant Mortality: _Plant size_  

Tests for plants 6 or more shoots dying from one year to the next. ***Note: ***These are not errors, these are plants whose size the year prior to being recorded as 'dead' in a survey was in the top 2% of dying plants.

**Test criteria**: 'warn' if $\geq$ 1 rows fail conditions, 'stop' if $\geq$ 2% of rows fail conditions.  


```{r agent9}
# check plnats
ha %>%
  filter(lag(census_status, 1) == "dead") %>% 
  create_agent(
    actions = al_default,
    label = "Check large plants dying"
  ) %>% 
  col_vals_lt(vars(shts), 6, na_pass = TRUE,
              label = "6 or more shoots") %>% 
  interrogate()
```



# **Return to the Github Repository:** [Link](https://github.com/BrunaLab/HeliconiaSurveys)


