---
title: "Data validation with the pointblank package"
author: Eric R. Scott
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

This uses `pointblank` to create a data validation report. In the resulting table at the end, any failing tests should have a CSV button that lets you download a .csv file of just the rows of data that don't pass that particular validation step.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(pointblank)
library(pointblank)
library(readr)
library(here)
library(dplyr)
library(stringr)
# Use `validate_rmd()` here to set options for the
# pointblank validation workflow within R Markdown documents
```


```{r load-data, message=FALSE, warning=FALSE}
ha <-
  read_csv(
    here("data_clean", "Ha_survey_with_Zombies.csv"),
    col_names = TRUE,
    cols(plot = col_character(),
         bdffp_reserve_no = col_character())
  )
```


**Action levels**

By default, warn if 1% of rows fail and error if 5% fail.  A stricter level warns if *any* rows fail, and errors if 10 or more rows fail.

```{r al, echo=TRUE}
al_default <-  action_levels(warn_at = 0.01, stop_at = 0.05)
al_strict <- action_levels(warn_at = 1, stop_at = 10)
```


```{r funs}
# Function for preconditions for some steps
calc_changes <- function(x) {
  x %>%
    group_by(HA_ID_Number) %>%
    arrange(HA_ID_Number, year) %>%
    mutate(ht_prev = lag(ht),
           shts_prev = lag(shts),
           ht_diff = ht - ht_prev,
           shts_diff = shts - shts_prev,
           ht_pc = abs(ht_diff / ht_prev),
           shts_pc = abs(shts_diff / shts_prev)) %>%
    ungroup()
}
```

```{r agent}
ha %>%
  create_agent(
    actions = al_default
  ) %>%
  col_is_character(vars(plot, habitat, ranch, bdffp_reserve_no, row, code.notes)) %>%
  col_is_numeric(vars(HA_ID_Number, tag_number ,column, year, ht, shts, infl)) %>%
  col_vals_in_set(vars(column), 1:10) %>%
  col_vals_in_set(vars(row), LETTERS[1:10]) %>%
  col_vals_in_set(vars(habitat), c("1-ha", "CF", "10-ha"),
                  actions = al_strict) %>%
  # check values are within reasonable range
  col_vals_between(vars(shts), 1, 20, na_pass = TRUE) %>%
  col_vals_between(vars(ht), 0, 200, na_pass = TRUE,
                   actions = al_strict) %>%
  col_vals_between(vars(infl), 0, 3, na_pass = TRUE) %>%
  # check that year-to-year change is reasonable
  col_vals_lt(vars(ht_pc), 3, na_pass = TRUE,
              preconditions = ~ . %>% calc_changes(),
              label = "% ht change < 300%") %>%
  col_vals_between(vars(ht_diff), -100, 100, na_pass = TRUE,
                   preconditions = ~. %>% calc_changes(),
                   label = "Ht change < 100cm",
                   actions = al_strict) %>%
  col_vals_between(vars(shts_diff), -5, 5, na_pass = TRUE,
                   preconditions = ~. %>% calc_changes(),
                   label = "shoot number change < 5",
                   actions = al_strict) %>%
  # check precision is correct
  col_vals_expr(label = "Height is measured to nearest cm",
                expr = ~ ht %% 1 == 0, actions = al_strict) %>%
  col_vals_expr(label = "Shoots is interger",
                expr = ~ shts %% 1 == 0, actions = al_strict) %>%
  col_vals_expr(label = "Number of inflorescences is integer",
                expr = ~ infl %% 1 == 0, actions = al_strict) %>%
  rows_distinct(actions = al_strict) %>%
  interrogate()
```
