---
title             : "**Demography of _Heliconia acuminata_ (Heliconiaceae) in an experimentally fragmented Amazonian landscape**"
shorttitle        : "Tropical plant demography"
author: 
    
      
  - name          : "Emilio M. Bruna"
    affiliation   : "1,2,3"
    email         : "embruna@ufl.edu"
    role:
      - Methodology
      - Data curation
      - Investigation
      - Funding acquisition
      - Conceptualization
      - Formal Analysis
      - Methodology
      - Project administration
      - Resources
      - Software
      - Supervision
      - Validation
      - Visualization
      - Writing – original draft
    corresponding : yes 
  - name          : "Maria Uriarte"
    affiliation   : "4"
    role:
      - Methodology
      - Investigation
      - Funding acquisition
      - Conceptualization
      - Formal Analysis
      - Methodology
      - Project administration
      - Resources
      - Software
      - Supervision
      - Validation
      - Visualization
      - Writing – review & editing
      
  - name          : "Maria Rosa Darrigo"
    affiliation   : "3"
    role:
      - Methodology
      - Investigation
      - Project administration
      - Writing – review & editing
  - name          : "Paulo Rubim"
    affiliation   : "3"
    role: 
      - Methodology
      - Investigation
      - Project administration
      - Writing – review & editing
  - name          : "Eric R. Scott"
    affiliation   : "1"
    role:
      - Methodology
      - Data curation
      - Software
      - Validation
      - Visualization
      - Writing – review & editing
  - name        : "W. John Kress"
    affiliation : "5"
    role:
      - Methodology
      - Investigation
      - Funding acquisition
      - Conceptualization
      - Methodology
      - Resources
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Department of Wildlife Ecology and Conservation, University of Florida, PO Box 110430, Gainesville, FL 32611-0430, USA"
  - id            : "2"
    institution   : "Center for Latin American Studies, University of Florida, PO Box 115530, Gainesville, FL 32611, USA"
  - id            : "3"
    institution   : "Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, AM 69011-970, Brazil"
  - id            : "4"
    institution   : "Department of Ecology, Evolution and Environmental Biology, Columbia University, 1200 Amsterdam Ave., New York, New York 10027, USA"
  - id            : "5"
    institution   : "Department of Botany, National Museum of Natural History, PO Box 37012, Smithsonian Institution, PO Box 37012, Washington DC, USA"
abstract: |
 Habitat fragmentation is thought to be a leading cause of extinction, but the demography of species in fragmented landscapes remains poorly understood. This is particularly true in tropical ecosystems, where studies monitoring populations of species in both fragments and areas of continuous habitat across all life-history stages are virtually nonexistent. Here we report 12 years (1997-2009) of annual censuses of 13 populations of the Amazonian understory herb _Heliconia acuminata_ (LC Rich.). These surveys were conducted in the experimentally fragmented landscape of the Biological Dynamics of Forest Fragments Project, located north of Manaus, Brazil. The plants are located in 50 x 100 m permanent plots located in forest fragments of different sizes (four 1-ha fragments and three 10-ha fragments) as well as 6 continuous forest sites. The population in each plot was censused annually, at which time we recorded the mortality of any previously marked plants and the size of surviving plants. We also marked and measured new seedlings. During the flowering season we also recorded the identity of flowering plants and how many inflorescences each of them produced. These data have been used in publications on topics ranging from how fragmentation-related reductions in germination influence population growth rates to tests of statistical methods for analyzing reproductive rates. 

Open Research Statement:


      
keywords: "Amazon, Brazil, deforestation, demography, edge effects, flowering, forest fragments, habitat fragmentation, Integral projection models, Matrix models, population dynamics, vital rates"
wordcount         : "------"
# bibliography      : ["r-references.bib"]
floatsintext      : no
figurelist        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : "papaja::apa6_pdf"
fig_caption: yes #yes will automatically add "Figure X" to the start 
always_allow_html: yes
header-includes:
  - \usepackage{booktabs}
  - \raggedbottom
 # - \usepackage{endfloat} #[nomarkers] excludes the {insert figure x around here] from main text. The others exclude the list of tables and figures. https://cs.brown.edu/about/system/managed/latex/doc/endfloat.pdf
  - \usepackage{tabu}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
# appendix: "Smith_etal_Appendix_QSSv1.Rmd"
csl: ecology.csl
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r GlobalOptions}
options(knitr.duplicate.label = 'allow')
```

```{r setup, include = FALSE}
# library("papaja")
# r_refs("APCpaper_QSS.bib")
knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE)
library(tidyverse)
library(gridExtra)
library(kableExtra)
# library(egg)
# library(magick)
# library(citr)
# library(devtools)
# devtools::install_github("crsh/papaja")
```


```{r cached=TRUE, echo = FALSE, warning=FALSE,message = FALSE}
# knitr::opts_chunk$set(echo = FALSE)
heliconia_data <- read_csv("./data_clean/HA_survey_pre_submission.csv", col_types = cols())

# Number of rows in the survey file
file_2_rows <- nrow(heliconia_data)+1 
file_2_size <- (file.info("./data_clean/HA_survey_pre_submission.csv")$size)/1000000
# heliconia_data$code

# No. of plants per plot
plants_per_plot <- heliconia_data %>% 
  select(HA_ID_Number, plot, plotID, habitat, bdffp_reserve_no, code) %>% 
  group_by(plotID) %>% 
  drop_na("plotID") %>% 
  summarise(count_plants=sum(n_distinct(HA_ID_Number)))

# SUMMARY STATISTICS

# Total no. of plants
total_plants <- plants_per_plot %>% 
  summarize(sum(count_plants))

# ULY plants
# From 1998-2009 there were -- -- -- such cases across all thirteen HDP plots. 
# Most were in the initial survey years (only XX% after 20XX) and 
# in high-density plots (XX% and XX% were in plots A and B, respectively).
# Most were in the initial survey years (only XX% after 20XX) and 
# in high-density plots (XX% and XX% were in plots A and B, respectively).

uly_per_plot <- heliconia_data %>% 
  select(HA_ID_Number, plot, plotID, habitat, bdffp_reserve_no, code) %>% 
  filter(code=="ULY (3)") %>% 
  group_by(plotID) %>% 
  summarise(count_uly=sum(n_distinct(HA_ID_Number)))

uly_per_yr_plot <- heliconia_data %>% 
  select(HA_ID_Number, year, plot, plotID, habitat, bdffp_reserve_no, code) %>% 
  # filter(year==1999) %>% 
  filter(code=="ULY (3)") %>% 
  group_by(plotID, year) %>% 
  summarise(count_uly=sum(n_distinct(HA_ID_Number))) %>% 
  arrange(desc(count_uly))

uly_table <- left_join(plants_per_plot,uly_per_plot) %>% 
  drop_na("plotID") %>% 
  mutate(perc=(count_uly/count_plants*100))




# Missing plants

# XX% of plants missing in a given year were found the following year
# XX% were eventually resurveyed. 
# XX% of plants marked missing were never remeasured. 
# This represents XX% and XX% of all plants in the database (i.e, missing plants are rare).


# Note this is the number of plamns that at some point in their history 
# were "missing" at least once
missing_per_plot <- heliconia_data %>% 
  select(HA_ID_Number, plot, plotID, habitat, bdffp_reserve_no, code) %>% 
  filter(code=="plant missing (60)") %>% 
  group_by(plotID) %>% 
  summarise(count_missing=sum(n_distinct(HA_ID_Number)))

missing_per_yr_plot <- heliconia_data %>% 
  select(HA_ID_Number, year, plot, plotID, habitat, bdffp_reserve_no, code) %>% 
  # filter(year==1999) %>% 
  filter(code=="plant missing (60)") %>% 
  group_by(plotID, year) %>% 
  summarise(count_missing=sum(n_distinct(HA_ID_Number))) %>% 
  arrange(desc(count_missing))

missing_table <- left_join(plants_per_plot,missing_per_plot) %>% 
  drop_na("plotID") %>% 
  mutate(perc=(count_missing/count_plants*100))
```

# INTRODUCTION

Understanding the consequences of habitat fragmentation remains a central area of ecological research [@didham2012; @driscoll2013; @selwood2015a; @brudvig2017; @resasco2017] decades after it was first identified as a threat to the integrity of ecosystems [@harris1984; @wilcove1986]. A large body of empirical research has documented myriad biotic changes associated with fragmentation, including the local extinction of plant species from fragments [@harrison1999; @laurance2011; @haddad2015; @fletcher2018]. Although the demographic mechanisms underlying these extinctions are rarely known [@bruna2009], they are often hypothesized to result from reduced rates of individual growth, reproduction, or survivorship in fragments [@zartman2015; @laurance1998]. This is especially true tropical forest fragments, where abiotic conditions can be dramatically different from those in primary forest [@broadbent2008]. Tests of this hypothesis remain limited, however, by the paucity of long-term and comprehensive data sets with which one can investigate the demography of tropical plants [@bruna2009].
 
\noindent 
**METADATA**

\noindent  
**CLASS I. DATA SET DESCRIPTORS**

\noindent  
**A. Data set identity:** Population structure and demography of the understory herb *Heliconia acuminata* (Heliconiaceae) in an experimentally fragmented landscape in the Central Amazon (1997-2009).

\noindent  
**B. Data set identification code:**
<!-- Database accession numbers or site-specific codes used to uniquely identify data set
-->
<!-- \setlength{\leftskip}{1cm} -->

1.  Data set 1: HDP_plot_descriptors.csv\
2.  Data set 2: HDP_data_1997-2009.csv\
3.  Data set 3: HDP_plot_maps.pdf

<!-- \setlength{\leftskip}{0pt} -->

\noindent 
**C. Data set description**


1. **Originators/Principal Investigator**: Emilio M. Bruna, Department of Wildlife Ecology and Conservation, University of Florida, PO Box 110430, Gainesville, FL 32611-0430, USA and Center for Latin American Studies, University of Florida, PO Box 115530, Gainesville, FL 32611, USA.

2. **Abstract**: The data set comprises 12 years (1997-2009) of annual censuses of 13 populations of the Amazonian understory herb *Heliconia acuminata* (LC Rich.). The data were collected at the Biological Dynamics of Forest Fragments Project, located north of Manaus, Brazil. The plants are located in 50 x 100 m permanent plots located in forest fragments of different sizes (four 1-ha fragments and three 10-ha fragments) as well as 6 continuous forest sites. The population in each plot was censused annually, at which time we recorded the mortality of any previously marked plants and the size of surviving plants. We also marked and measured new seedlings. During the flowering season we also recorded the identity of flowering plants and how many inflorescences each of them produced. The data set can be used to study the ecological factors influencing the demography and population dynamics of tropical plants, as well as the consequences for plant demography and population dynamics of anthropogenic impacts such as deforestation or climate change. 

\noindent  
D. **Key words:** Amazon, Brazil, deforestation, demography, edge effects, flowering, forest fragments, habitat fragmentation, integral projection models, matrix models, population dynamics, vital rates.

\noindent 
**CLASS II. RESEARCH ORIGIN DESCRIPTORS**

\noindent  
**A. Overall project description:**

1.  **Identity:** The *Heliconia* Demography Project (HDP)

2.  **Originators:** Emilio M. Bruna, W. John Kress, and María Uriarte

3.  **Period of study:** 1997-2009

4.  **Objectives:** To understand how habitat fragmentation influences the demography and population dynamics of the Amazonian understory herb *Heliconia acuminata* (Heliconiaceae) using experiments, observation data, and demographic modeling. The HDP is based at the Biological Dynamics of Forest Fragments Project, located north of Manaus, Amazonas, Brazil

5.  **Sources of funding:** The initial establishment of plots  and 1998-2002 surveys were supported by awards to E. M. Bruna from the Smithsonian Institution (Graduate Student Research Award), the University of California, Davis (Center for Population Biology and M. E. Mathias Graduate Research Grants), the Biological Dynamics of Forest Fragments Project (Graduate Student Logistics Grant), the National Science Foundation (Dissertation Improvement Grant INT 98-06351), and the Ford Foundation (Dissertation Year Fellowship). The 2001-2005 surveys were supported a grant from the National Science Foundation to E. M. Bruna (Research Starter Grant DEB-0309819). The 2006-2009 surveys were supported by grants from the National Science Foundation to E. M. Bruna (DEB-0614149) and María Uriarte (DEB-0614339).

\noindent  
**B. Subproject description:** The core of the *Heliconia* Demography Project (HDP) is annual censuses of thirteen *H. acuminata* populations located in either continuous forest or one of the BDFFP's experimentally isolated forest fragments. In addition to recording the reproduction, growth, and survivorship of established plants, newly recruited seedlings are marked, measured, and assigned their own unique identification number. These data can be used for a broad range of analyses, but the primary purpose behind their collection was to parameterize size-structured demographic models [@caswell_matrix_2000; @ellner_integral_2006] with which to compare the demography and population dynamics of *H. acuminata* populations in fragments and continuous forest. This paper reports 12 years of annual census data (1997-2009)

1.  **Site description**

<!-- -->

a.  **Study Site:** The data were collected at the Biological Dynamics of Forest Fragments Project (BDFFP), located approximately 70 km north of Manaus, Amazonas, Brazil (2°30'S, 60°W, **Fig. X**). The BDFFP is a 1000-km mosaic of lowland forest, forest fragments, secondary forests, and pastures [@lessons2002]. It is currently administered collaboratively by the Smithsonian Tropical Research Institute and Brazil's Instituto Nacional de Pesquisas da Amazônia (INPA).

> The BDFFP landscape includes catchments of the Urubu, Cuieiras, and Preto da Eva rivers [@nessimian2008]; the often rugged topography ranges in elevation from 50-150 m elevation [@gascon2001. The forest canopy in the sites is \~30--35 m tall, with emergent trees of up to \~55 m. The tree community at the BDFFP is highly diverse [~1300 species;  @laurance_hyper-diverse_2001], with as many as 280 tree species ha$^-$$^1$ [@de_oliveira_central_1999]. The understory is dominated by stemless palms [@scariot_forest_1999]. Soils in the sites are nutrient-poor xanthic ferralsols, known as yellow latosols in the Brazilian soil classification system. Despite their high clay content they have poor water-retention capacity  [@fearnside2001]. All demographic plots are located in non-flooded (i.e., terra firme) forest; none are bisected by streams.

b.  **Climate:** Mean annual temperature at the site is 26$^\circ$C (range 19-39$^\circ$C). Annual rainfall ranges from 1900-2300 mm **[scott gcbn paper]**, with a pronounced dry season from June-December in which there is \<100 mm rain per month. 

c.  **Site history:** A complete history of the BDFFP can be found in @gascon_study_2001. Briefly, the BDFFP fragments were created between 1980-1984 by felling the trees surrounding the forest to be isolated [@lovejoy1986]. Fragments were fenced to prevent incursion by cattle from the surrounding pastures. To ensure fragments remain isolated, a 100m strip around each fragment is regularly cleared of the secondary growth [@gascon_study_2001]. The structure and species composition of the secondary growth that surrounds a fragment, which is strongly dependent on whether fire was used to clear land prior to planting pasture grasses [@mesquita2001], can have large effects on the biological dynamics and abiotic conditions in fragments [@laurance2002; @laurance2011a].

<!-- -->

2.  **Sampling Design**

<!-- -->

a.  **Focal species:** *Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot native to Amazonia [@kress1990] and widely distributed throughout the Amazon basin [@kress_diversity_1990]. Although many species of *Heliconia* grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others, including *H. acuminata*, grow primarily in the shaded forest understory [@kress1983]. It is the most abundant understory herb throughout much of the BDFFP [@ribeiro2010a].

> Although *Heliconia acuminata* can be propagated by segmenting the rhizome [@bruna_edge_2011], vegetative reproduction in the field is limited and recruitment primarily via seed [@bruna1999; @bruna2002b]. *Heliconia acuminata* flowers during the rainy season, when reproductive plants produce 1 or more bracts subtending up to 25 **[mean?range?]** flowers each [@bruna2002]. Flowers are pollinated by the 'traplining' hummingbirds *Phaeothornis superciliosus* and *P. bourcieri* [@bruna2004]. Flowers are open only 1 day and visitation rates are low (<1 visit per hour, [@bruna2004]. Fruits produced by successfully pollinated flowers have up to three seeds and are dispersed by birds [@uriarte2011]. Seeds germinate 6-7 months after dispersal at the onset of the rainy season [@bruna1999; @bruna2002b].

> *Heliconia acuminata* leaves are ocassionally damaged by the hispine beetle *Cephaloleia nigriceps* Baly **[pers comm]**, but the amount of leaf tissue removed is minimal [@bruna2002a; @bruna2005]. However, these beetles can cause extensive damage to the bracts, ovaries, and developing fruits. Experiments also indicate post-dispersal seed predation is very low and that even seeds protected from burial under leaf-litter rarely germinate after one year, suggesting *H. acuminata* has no seed bank [@bruna1999; @bruna2002b]; the lack of a seed bank could be due to limited seed viability, mortality resulting from burial under leaf-litter, or both [@bruna2005].

a.  **Taxonomy, systematics, and voucher specimens:** *Heliconia* is the only genus in the family Heliconiaceae. This family is distinguished from the others in the order Zingiberales by inverted flowers, having a single staminode, and fruits that are drupes. It is estimated that there are 200-250 species of *Heliconia*, almost all of which are native to the Neotropics. *Heliconia acuminata* L. C. (Rich.) (Richard 1831) is one of the approximately 20 *Heliconia* species found in the Brazilian Amazon [@kress1990a].

> Voucher specimens of *H. acuminata* collected at the BDFFP as part of this study can be found at the herbaria of the University of California Davis (accession numbers DAV 69391-69396) and the Instituto Nacional de Pesquisas da Amazônia (accession numbers INPA 189569-189573).

b.  **Study Design:** The study consists of 13 demographic plots distributed across the BDFFP landscape (Bruna and Kress 2002). Six plots are in continuous forest, four are in 1-ha fragments, and three are in 10-ha fragments. Each demographic plot is 50m x 100m and is subdivided into 50 contiguous subplots (each 10 x 10 m) to facilitate the surveys. Plots in 1-ha fragments were established in a randomly selected half of the fragment, plots in 10-ha fragments are in the center of the fragment, and plots in continuous forest sites are located 500-4000 m from any the borders of primary forest with secondary forest or pastures. The plots furthest apart are separated by \~65 km.

> Plots 1-ha fragments, 10-ha fragments, and three of the continuous forest sites were established from January-April 1997, the remaining three plots in continuous forest were established in January 1998. To mark the plants, a team of 2-3 people slowly walked through each subplot and located all *Heliconia acuminata* and marked them with a wooden stake to which was attached an individually numbered aluminum tag (Racetrack Aluminum Tags, Forestry Suppliers). The size of each plant was measured in two ways: (1) by counting the number of vegetative shoots it had and (2) by measuring the height of the plant from the ground to the top of its highest leaf (to the nearest cm using a tape measure). Three additional plots were established in continuous forest sites in 1998 (CF 4-6); all plants in these plots were tagged and measuring in the same way as in other plots.>

>Plots were censused annually with the onset of the rainy season and seedling establishment (generally late January to February). The exception to this was the three continuous forest plots establis>hed in August 1998, which were censused in August 1999. During each census, team members recorded the size of all surviving plants, marked and measured all new seedlings, and measured all surviving plants. Regular visits were made to all 13 plots throughout the rainy season to identify reproductive individuals, at which time we recorded the number of inflorescences.

3.  **Project personnel:** In addition to the Principal Investigator and Originators, other key personnel include the Technicians that were responsible for coordinating the annual censuses, the BDFFP field staff (*"Mateiros"*) that assisted with data collection and provided support in the field, and student or postgraduate field assistants hired to assist with the surveys.

a.  **Technicians:** Paulo Rubim (2007-2012), Maria Beatriz Nogueira (2002), Maria Rosa Darrigo (2002-2003), Cris Follman Jurinitz (2003), Simone Benedet (2004).

b.  **BDFFP Staff:** Osmaildo Ferreira da Silva, Francisco Marques, Alaercio Marajó dos Reis, João de Deus Fragata, Romeu Cardoso.

c.  **Student & Postgraduate Field Assistants:** Olavo Nardy (2000), Obed Garcia (2001), Sylvia Heredia (2001-2002), Maria Beatriz Nogueira (2002), Cris Follman Jurinitz (2003), David M. Lapola (2003), Denise Cruz (2003), Cristina Escate (2004), Bruno Turbiani (2005), Elisabete Marques da Costa (2006), Wesley Dáttilo da Cruz (2007), Jefferson José Valsko da Silva (2007). 

\noindent  
**CLASS III. DATA SET STATUS AND ACCESSIBILITY**

\noindent  
**A. Status**

1.  **Latest update:** `r format(Sys.time(), '%d %B %Y')`

2.  **Latest archive date:** `r format(Sys.time(), '%d %B %Y')`

3.  **Metadata status:** Complete and up to date (`r format(Sys.time(), '%d %B %Y')`)

4.  **Data verification & Quality Control Procedures:** Data verification and quality control is ongoing. After each survey, the measurements for a plant are compared with those from previous years to identify outliers that could represent potential errors in either the recording of measurements in the field or data entry (e.g., a plant with 1 shoot in year N and 11 shoots in year N+1 is likely an error in data entry). Discrepancies are verified by first double-checking data sheets and, if necessary, returning to the field to remeasure plants. Occasionally the quadrat in which plants are located might change from one survey to the next (i.e., a plant on the edge of plots A1 and A2 might be recorded in A1 in year N but the survey team decided it is in A2 in year N+1. In these cases, the most current location is used unless a subsequent survey indicates otherwise.

a. **Established plants not recorded in prior surveys:** Occasionally unmarked post-seedling plants will be found in a plot. If all other previously marked plants in the plot have been located, this indicates previous survey teams failed to find and record the plant. Such plants are marked with a new numbered tag, mapped, measured, and added to the database for measurement in subsequent years. Their record in the database for the year they were discovered includes a code identifying them as a mature plant not recorded in previous surveys (see Section IV, Table 2). From 1998-2009 there were **-- -- --** such cases across all thirteen HDP plots. Most were in the initial survey years (only **XX%** after **20XX**) and in high-density plots (**XX% and XX%** were in plots A and B, respectively).

b. **Tag loss:** Treefalls and other disturbances sometimes cause the stakes to which numbered tags are attached to be displaced, broken, or buried under leaf litter. If after an extensive search the tag still can't be found, the plant is marked with a new stake and numbered tag and a notation is made in the survey record indicating a new number was
given to a plant with a lost tag. In some cases, determining the identify of a plant missing its tag is straightforward (e.g., all plants in a plot are found except one, and the plant without a tag is similar in height and shoots number as the plant that was missing). In those cases, the plant's ID number is updated in the database and the change is logged. In some cases, however, it may not be possible to definitely conclude with which number a plant was originally marked (e.g., when two adjacent and similarly sized plants are both missing their tags). In such cases we review the growth history of the plants in question to decide which one should be assigned each new number. The change in number is again logged; a list of all changes in tag number to date is permanently archived at [zenodo] and updated at [github site]. From 1998-2009 there were **-- -- --** such cases across all thirteen HDP plots. Most were in the in plots A,B, and C **XX%, XX% and XX%**, respectively).

c. **Plants under treefalls:** Plants trapped under the large crowns of fallen trees may go several years without being measured until leaves have dropped and the area under the crown can be safely searched. In some cases, neither the plant nor the ID tag can be found. When this occurs, the plant is recorded as missing but not removed from the record. (see Section IV, Table 2 for the code added to the record in such cases). **XX%** of plants missing in a given year were found the following year and XX% were eventually resurveyed. **XX%** of plants marked missing were never remeasured. This represents **XX% and XX%** of all plants in the database (i.e, missing plants are rare).

d. **Pre-archiving review:** An extensive review of the data and quality control effort was conducted prior to the publication of the dataset. Any possible errors (e.g. plants recorded as dead with a measurement in a subsequent year, duplicated tag numbers, doubts regarding the quadrat in which a plant is located) were investigated E. M. Bruna using the original datasheets and electronic data entry files (both stored at the University of Florida). All resulting corrections were added to the database and logged. The log of changes to date is archived at [zenodo]; updates will be posted at [github site]. Questions from future users of the database should be referred to E. M. Bruna, who will investigate and update the database as needed.

\noindent 
**B. Accessibility**

1. **Storage location and medium:** Ecological Society of America data archives [Ecological Archives URL].
    
2. **Location of original data forms, electronic files, and archived copies:** Original data sheets and electronic data files are with E. M. Bruna at the University of Florida. Original data files and paper copies are stored in separate campus locations, with electronic copies (.pdf format) stored on a desktop computer, portable hard drive, and a cloud storage account. The electronic files into which each year's survey data were entered, and a copy of the complete database, are stored on a desktop computer with copies on a portable hard drive and UF cloud storage account.

3.  **Contact person(s):** Emilio M. Bruna, Department of Wildlife Ecology and Conservation, Box 110430, Gainesville, FL 32611 USA. Phone: (352) 846-0634. Email: [embruna\@ufl.edu](mailto:embruna@ufl.edu){.email}

4.  **Copyright restrictions:** None

5.  **Proprietary restrictions:** None. However, we do request that authors of publications using these data (1) cite this data paper as per Ecological Archives policy, (2) register the publication as part of the BDFFP Technical Series by contacting the BDFFP Director or E. M. Bruna, and (3) provide E. M. Bruna with a copy of their article upon acceptance. This allows us to track the data set's usage, advise users of any corrections, report articles using the data to the funding agencies that provided support, and document that different ways in which the scientific community uses the data.

\noindent  
**CLASS IV. DATA STRUCTURAL DESCRIPTORS**

\noindent  
**A. Data set file:** Descriptors of the *Heliconia* demographic plots

1.  **Identity:** HDP_plot_descriptors.csv

2.  **Size:** ----- rows (including header), --- kilobytes.

3.  **Format and storage mode:** ASCII text, comma delimited. No
    compression scheme used.

4.  **Header information:** The first row of the file contains the variable names described in Table 1 below.

5.  **Alphanumeric attributes:** Mixed.

6.  **Missing value codes:** Missing values are represented with NA.

7.  **Data anomalies:** See Section IV, Table 1 for codes used to denote modified data or record unique or unusual circumstances regarding a record.

8.  **Variable information:** Each row in the data set is a demographic plot, with columns of data describing that plot. Blanks do not denote missing information, but rather nothing relevant to report.

# [INSERT TABLE 1 HERE]

\noindent 
**B. Data set file:** *Heliconia* Demographic Data

1.  **Identity:** HDP_data_1997-2009.csv

2.  **Size:** `r file_2_rows` rows (including header), `r file_2_size`
    kilobytes.

3.  **Format and storage mode:** ASCII text, comma delimited. No compression scheme used.

4.  **Header information:** The first row of the file contains the variable names in Table 2 below.

5.  **Alphanumeric attributes:** Mixed.

6.  **Missing value codes:** Missing values are represented with NA.

7.  **Data anomalies:** See Section IV, Table 2 for codes used to denote modified data or record unique or unusual circumstances regarding a record, etc.

8.  **Variable information:** Each row in the dataset is the data collected on an individual plant. For notes, blanks do not denote missing information. Instead, they denote nothing relevant to report.

9.  **Computer programs and data-processing algorithms:** code to append the plot descriptors can be found at Zenodo. This will be updated regularly, between updates the code can be found at gtihub HDP,

<!-- Table 2: Description of the column names for HDP_data_1997-2009.csv and description of the data in each column -->

# [INSERT TABLE 2 HERE]

\noindent  
**C. Data set file:** Maps of the HDP plots

1.  **Identity:** HDP_plot_maps.pdf

2.  **Size:** --- kilobytes.

3.  **Format and storage mode:** pdf file. No compression scheme used.

4.  **Description:** maps of the HDP plots, their location and orientation in the BDFFP reserves, and the numbering of subplots used to map plants.

\noindent  
**CLASS V. SUPPLEMENTAL DESCRIPTORS**

\noindent  
**A. Publications and results:** The following list includes articles to date that used part or all of the dataset in their analyses. An update list can be found at [github].

1.  Bruna, E. M. and W. J. Kress. 2002. Habitat fragmentation and the demographic structure of an Amazonian understory herb (*Heliconia acuminata*). *Conservation Biology*, 16(5): 1256-1266.

2.  Bruna, E. M., O. Nardy, S. Y. Strauss, and S. P. Harrison. 2002. Experimental assessment of *Heliconia acuminata* growth in a fragmented Amazonian landscape. *Journal of Ecology*, 90(4): 639-649.

3.  Bruna, E. M. 2002. Effects of forest fragmentation on *Heliconia acuminata* seedling recruitment in the central Amazon. *Oecologia*, 132:235-243.

4.  Bruna, E. M. 2003. Are plant populations in fragmented habitats recruitment limited? Tests with an Amazonian herb. *Ecology*, 84(4): 932-947.

5.  Bruna, E. M. 2004. Biological impacts of deforestation and fragmentation. Pages 85-90 in *The Encyclopaedia of Forest Sciences*. J. Burley, J Evans, and J Youngquist, (eds.). Elsevier Press, London.

6.  Morris, W. F., C. A. Pfister, S. Tuljapurkar, C. V. Haridas, C. Boggs, M. S. Boyce, E. M. Bruna, D. R. Church, T. Coulson, D. F. Doak,, S. Forsyth, J-M. Gaillard, C. C. Horvitz, S. Kalisz, B. E. Kendall, T. M. Knight, C. T. Lee, E. S. Menges. 2008. Longevity can buffer plant and animal populations against changing climatic variability. *Ecology* 89(1): 19-25.

7.  Fiske, I., E. M. Bruna, and B. M. Bolker. 2008. Effect of sample size on estimates of population growth rates calculated with matrix models. *PLoS ONE* 3(8): e3080.

8.  Fiske, I. and E. M. Bruna. 2010. Alternative spatial sampling in studies of plant demography: consequences for estimates of population growth rate. *Plant Ecology* 207(2) 213-225.

9.  Uriarte, M., E. M. Bruna, P. Rubim, M. Anciaes, and I. Jonckeeere. 2010. Effects of forest fragmentation on seedling recruitment of an understory herb: assessing seed vs. safe-site limitation. *Ecology* 91(5):1317-1328.

10. Gagnon, P. R., E. M. Bruna, P. Rubim, M. R. Darrigo, R. C. Littlel, M. Uriarte, and W. J. Kress. 2011. The growth of an understory herb is chronically reduced in Amazonian forest fragments. *Biological Conservation* 144: 830-835.

11. Uriarte, M. Anciães, M. T.B. da Silva, P. Rubim, E. Johnson, and E. M. Bruna. 2011. Disentangling the drivers of reduced long-distance seed dispersal by birds in an experimentally fragmented landscape. *Ecology* 92(4): 924-93.

12. Côrtes, M., M. Uriarte, M. Lemes, R. Gribel, W. J. Kress, P. Smouse, E. M. Bruna. 2013. Low plant density enhances gene flow in the Amazonian understory herb *Heliconia acuminata*. *Molecular Ecology* 22: 5716-5729.

13. Brooks, M. E., K. Kristensen, M. R. Darrigo, P. Rubim, M. Uriarte, E. M. Bruna, B. M Bolker. 2019. Statistical modeling of patterns in annual reproductive rates. *Ecology* 100(7): e02706.
  
14. Scott, E. R., M. Uriarte, and E. M. Bruna. 2022. Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments. _Global Change Biology_. 28(2):463-479.

# ACKNOWLEDGMENTS

This work would not have been possible without the many technicians and field assistants participating in the annual surveys, the logistical support provided by the staff of the BDFFP, and the organizations that provided financial support. Sharon Y. Strauss, Susan P. Harrison, and Patricia Delamônica Sampaio provided invaluable support and feedback during the project's initial phase. This paper is publication number -- -- -- in the BDFFP Technical Series.

<!-- Tables, figures, and appendices. Tables and figures should be embedded in the metadata where appropriate. Tables should be in HTML and figures should be embedded .JPG, .GIF, or .PNG files. Appendices are not acceptable parts of data papers. -->

\newpage

# References

```{=tex}
\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}
```
::: {#refs custom-style="Bibliography"}
:::

\endgroup
\newpage

# FIGURES
1. satellite image
2. population structure
3. plot maps
\newpage

```{r Table1, echo=FALSE, message = FALSE, warning=FALSE}
Table1 <- read_csv("./Manuscript/table_1.csv", col_names = TRUE) %>% 
  # rename("Storage type"="Storage","List and definition of variable codes"="Codes") %>% 
  # replace_na(list("Variable" = "",	"Definition" = "",	"Storage type" = "",	"List and definition of variable codes" = "",	"Units, range, and precision"=""))
replace_na(list("Variable" = "",	"Definition" = "",	"Storage type" = "",	"List and definition of variable codes" = ""))
Table1<-as.data.frame(Table1)
# Table1_names <- c("Variable",	"Definition",	"Storage type",	"List & definition of variable codes",	"Units, range, & precision") 

tmp <- knitr::kable(Table1,
     digits = 2,
     format = "latex",
     align="l",
     # escape=FALSE,
     row.names = FALSE,
     booktabs=T,
    linesep = "", #removes the blank line after every 5 lines
     caption = "Description of the column names for Heliconia-plot-descriptors.csv and description of the data in each column") %>%
   kable_styling(bootstrap_options = c("hover"),
                 full_width = TRUE,
                 # latex_options = c("scale_down"),
                 font_size = 10,
                 position = "center")
    # ) %>%
    # column_spec(2, width = "10em")
  # row_spec(2, extra_latex_after = "\\cline{1-5}") %>%
  # row_spec(15, extra_latex_after = "\\cline{1-5}") %>%
  # row_spec(25, extra_latex_after = "\\cline{1-5}") %>%
  # row_spec(32, extra_latex_after = "\\cline{1-5}") %>%
  # row_spec(42, extra_latex_after = "\\cline{1-5}") %>%
  # row_spec(55, extra_latex_after = "\\cline{1-5}")
knitr::asis_output(tmp)
```

```{r Table2, echo=FALSE, message = FALSE, warning=FALSE}
Table2 <- read_csv("./Manuscript/table_2.csv", col_names = TRUE) %>% 
  replace_na(list("Variable" = "",	"Definition" = "",	"Storage type" = "",	"List and definition of variable codes" = "",	"Units, range, and precision"=""))
Table1<-as.data.frame(Table2)
# Table1_names <- c("Variable",	"Definition",	"Storage type",	"List & definition of variable codes",	"Units, range, & precision") 

tmp <- knitr::kable(Table2,
     digits = 2,
     format = "latex",
     align="l",
     # escape=FALSE,
     row.names = FALSE,
     booktabs=T,
    linesep = "", #removes the blank line after every 5 lines
     caption = "Description of the column names for HDP-data-1997-2009.csv and description of the data in each column") %>%
   kable_styling(bootstrap_options = c("hover"),
                 full_width = TRUE,
                 latex_options = c("scale_down"),
                 font_size = 10,
                 position = "center")
knitr::asis_output(tmp)
```

