---
title: "Tag Change Records"
# date: "updated: `r Sys.Date()`"
# layout: page
# permalink: HeliconiaSurveys/TagChanges
# header-includes:
# - \AtBeginDocument{\let\maketitle\relax} # this removes default title section
# - \usepackage[default]{sourcesanspro}
# - \usepackage{fancyhdr} # modify header into left, center, right
# - \pagestyle{fancy} # cues the use of fancy headers
# - \fancyfoot{} # cues the use of fancy footer
# - \fancyhead[R]{HDP Tag Changes, p. \thepage} 
# - \fancyfoot[R]{Last update `r format(Sys.time(), '%d %B %Y')`}
# 
# - \usepackage{parskip} # manipulate indents and spacing before/after paragra phs 
# - \usepackage{xcolor} # changes colors of text
# # more info # https://www.ctan.org/pkg/sectsty
# - \usepackage{sectsty} \sectionfont{\centering\color{darkmidnightblue}}
# - \usepackage{sectsty} \subsectionfont{\color{darkmidnightblue}}
# # http://latexcolor.com/ to choose colors
# - \definecolor{darkcerulean}{rgb}{0.03, 0.27, 0.49}
# - \definecolor{darkmidnightblue}{rgb}{0.0, 0.2, 0.4}
# # https://tex.stackexchange.com/questions/258714/prevent-hyphenation-across-lines-enforce-right-margin
# # https://babichmorrowc.github.io/post/changing-fonts-in-rmarkdown-pdfs/
# # bibliography: references.bib
# mainfont: SourceSansPro
# geometry: margin=1in
# fontsize: 12pt
# linkcolor: blue
# urlcolor: blue
# output: html_document
output: 
  rmdformats::robobook
#   # rmdformats::downcute:
    
self_contained: true
thumbnails: true
lightbox: true
gallery: false
highlight: tango
bookdown: true
---



```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
library(tidyverse)
library(here)
library(kableExtra)
here::here()
knitr::opts_chunk$set(echo = TRUE)
## Global options
options(max.print="75")

opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r data, cached=TRUE, echo = FALSE, warning=FALSE,message = FALSE}

tag_changes <- read_csv(here("data","survey_raw","tag_changes.csv"), col_types = cols())
ha_sites <- read_csv(here("data","survey_clean","heliconia_survey_clean.csv"), col_types = cols()) %>% 
  select(habitat,plot_id,plot,bdffp_reserve_no,ranch) %>% 
  unique()
tag_changes<-left_join(tag_changes,ha_sites, by="plot") %>% 
  select(habitat,ranch,plot_id,plot,year,old_tag_no,new_tag_number,notes)
rm(ha_sites)

# Standardize 

tag_changes<-tag_changes %>% 
  mutate(ranch = 
           case_when(
             ranch == "PortoAlegre" ~ "PAL",
             .default = as.character(ranch)
             )
         ) %>% 
  rename(new_tag_no=new_tag_number)


# ANY CHANGES TO THE TAG NUMBERS AS OF June 1, 2023 ARE ENTERED MANUALLY HERE

# 2024 Minisurvey done to prep for grant proposal

tag_changes_2024_CF2 <- data.frame(old_tag_no=character(),
                                   new_tag_no=integer(),
                                   notes=character(),
                                   stringsAsFactors=FALSE)
tag_changes_2024_CF2 <- tag_changes_2024_CF2 %>% 
  add_row(
          old_tag_no="153",
          new_tag_no=13902,
          notes=NA
          ) %>% 
  add_row(old_tag_no="2361",
          new_tag_no=13908,
          notes=NA
          ) %>% 
  add_row(old_tag_no="1340",
          new_tag_no=13907,
          notes=NA
          ) %>% 
  add_row(old_tag_no="2365",
          new_tag_no=13909,
          notes=NA
          ) %>% 
     add_row(old_tag_no="2390",
             new_tag_no=13560,
             notes=NA
          ) %>% 
  add_row(old_tag_no="767",
          new_tag_no=13910,
          notes=NA
          ) %>% 
 add_row(old_tag_no="147",
          new_tag_no=13901,
          notes=NA
          ) %>% 
  add_row(old_tag_no="460",
          new_tag_no=13905,
          notes=NA
          ) %>% 
  add_row(old_tag_no="461",
          new_tag_no=13904,
          notes=NA
          ) %>% 
  mutate(habitat="CF",
          ranch="EST",
          plot_id="CF-2",
          plot="5750",
          year=2024) 

# Add back to file, rearrange, and save
tag_changes_updated<-bind_rows(tag_changes,tag_changes_2024_CF2)

# Arrange and save updated version

tag_changes_updated<-tag_changes_updated %>%
  arrange(habitat,
          ranch,
          plot_id,
          year,
          old_tag_no)

write_csv(tag_changes_updated, here("data","survey_clean","tag_changes.csv"))
```


<!-- ## Tag Number Changes -->

**Last update:** `r format(Sys.time(), '%d %B %Y')`

Field survey teams periodically change the tags attached to individual plants (e.g., if the tag becomes difficult to read or the tag is lost under a fallen branch). Below is (a) a summary of the number of tags changed per plot and (b) a list of the individual changes; these records are necessary if using the original data sheets to validate plant records.    

## Summary of tag replacements.



```{r nplants_summary, cached=TRUE, warning=FALSE,message = FALSE}

total_tagged <- read_csv(here("data","survey_clean","heliconia_survey_clean.csv"), col_types = cols()) %>% 
  group_by(habitat,plot_id) %>% 
  summarise(total_tagged=n_distinct(tag_number))

tag_change_summary<-tag_changes_updated %>% 
  group_by(habitat,plot_id,ranch) %>% 
  tally() %>% 
  left_join(total_tagged) %>% 
  rename(tags_changed=n) %>% 
  ungroup()
tag_change_summary<-tag_change_summary %>% 
  mutate(percent_changed=tags_changed/total_tagged*100) %>% 
  mutate(percent_changed=round(percent_changed,2)) %>% 
  select(-habitat) %>% 
    arrange(ranch,
          plot_id
          )



kbl(tag_change_summary,
    align = "rrrrr") %>% 
pack_rows("1-ha Fragments", 1, 4) %>% 
  pack_rows("10-ha Fragments", 5, 7) %>% 
  pack_rows("Continuous Forest", 8, 13) %>%
  column_spec(1, width = "15em")



```

---

## List of all tags replaced.

```{r nplants_list, cached=TRUE, warning=FALSE,message = FALSE}


options(knitr.kable.NA = '')

kbl(tag_changes_updated,
    align = "cccccccc")


```


