---
title: "Data validation with the pointblank package"
author: Eric R. Scott
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

This uses `pointblank` to create a data validation report. In the resulting table at the end, any failing tests should have a CSV button that lets you download a .csv file of just the rows of data that don't pass that particular validation step.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

library(pointblank)
library(readr)
library(here)
library(dplyr)
library(stringr)
library(ggplot2)
library(gghighlight)
# Use `validate_rmd()` here to set options for the
# pointblank validation workflow within R Markdown documents
```


```{r load-data}
ha <- read_csv(
  here("data", "data_archive", "HDP_1998_2009.csv")
)
ha_plots <- read_csv(
  here("data", "data_archive", "HDP_plots.csv")
)
```

# Check for missing values: Height

```{r echo=FALSE}
ha %>% 
  group_by(plot_id, year) %>% 
  summarize(prop_na = sum(is.na(ht))/n(), .groups = "drop") %>% 
  ggplot(aes(x = year, y = prop_na)) +
  geom_point(color="red") +
  facet_wrap(~plot_id) +
  theme_bw() +
  scale_x_continuous(breaks = scales::breaks_pretty()) +
  scale_y_continuous(breaks = c(0,0.5,1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  gghighlight(prop_na > .5) +
  labs(y = "Proportion of values missing - Height")
  
```

# Check for missing values: Shoots
```{r echo=FALSE}
ha %>% 
  group_by(plot_id, year) %>% 
  summarize(prop_na = sum(is.na(shts))/n(), .groups = "drop") %>% 
  ggplot(aes(x = year, y = prop_na)) +
  geom_point(color="red") +
  facet_wrap(~plot_id) +
  theme_bw() +
  scale_x_continuous(breaks = scales::breaks_pretty()) +
  scale_y_continuous(breaks = c(0,0.5,1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  gghighlight(prop_na > .5) +
  labs(y = "Proportion values missing - Shoots")
  
```



**Action levels**

By default, warn if 1 or more rows fail conditions and error if 2% or more fail.  Some checks are run with a stricter action level that errors if *any* rows fail.

```{r al, echo=TRUE}
al_default <-  action_levels(warn_at = 1, stop_at = 0.02) #warn if even row fails, error if 2% of rows fail
al_strict <- action_levels(stop_at = 1) #error if even one row fails
```


```{r funs}
#calculate percent changes
ha_change <- 
  ha %>%
  group_by(plant_id) %>%
  arrange(plant_id, year) %>%
  mutate(ht_prev = lag(ht),
         shts_prev = lag(shts),
         ht_diff = ht - ht_prev,
         shts_diff = shts - shts_prev,
         ht_pc = abs(ht_diff / ht_prev),
         shts_pc = abs(shts_diff / shts_prev)) %>%
  #re-order columns so easier to see errors
  select(plot_id, subplot, plant_id, year, shts_prev, shts, shts_diff, shts_pc, ht_prev, ht, ht_diff, ht_pc, everything()) %>% 
  ungroup()
```

# Data Validation

## These are evaluations of (1) data type, (2) the range of values for shoots, height, and infloresences, and (3) potential duplicates in `HDP_plots.csv` and `HDP_1997_2009.csv`.


```{r agent1}
ha %>%
  create_agent(
    actions = al_default,
    label = "Data Validation"
  ) %>%
  col_vals_in_set(vars(subplot), unlist(lapply(LETTERS[1:10], paste0, 1:10)),
                  actions = al_strict) %>%
  col_vals_in_set(vars(plot_id), ha_plots$plot_id,
                  actions = al_strict) %>%
  col_vals_expr(label = "Height is measured to nearest cm",
                expr = ~ ht %% 1 == 0, actions = al_strict) %>%
  col_vals_expr(label = "Shoots is interger",
                expr = ~ shts %% 1 == 0, actions = al_strict) %>%
  col_vals_expr(label = "Number of inflorescences is integer",
                expr = ~ infl %% 1 == 0, actions = al_strict) %>% 
  # check values are within reasonable range
  col_vals_between(vars(shts), 0, 20, na_pass = TRUE,
                   label = "shoots between 0 and 20") %>%
  col_vals_between(vars(ht), 0, 200, na_pass = TRUE,
                   label = "height between 0 and 200cm") %>%
  col_vals_between(vars(infl), 0, 3, na_pass = TRUE,
                   label = "infloresences between 0 and 3") %>%
  rows_distinct(actions = al_strict, 
                label  = "duplicated rows") %>% 
  col_vals_not_null(vars(plant_id), actions = al_strict) %>% 
  rows_distinct(columns = "plant_id",
                segments = vars(year),
                actions = al_strict,
                label = "Check for duplicate ID's within each year") %>% 
  interrogate()
```


# Year to year change

## This test is for unusually large changes in plant size from year t to t+1. 

Checks that year to year change in size is reasonable

```{r agent2}
ha_change %>%
  create_agent(
    actions = al_default,
    label = "Check growth & regression"
  ) %>%
  
  # check that year-to-year change is reasonable
  col_vals_lt(vars(ht_pc), 2, na_pass = TRUE,
              label = "|% change in height| < 200%") %>%
  col_vals_between(vars(ht_diff), -100, 100, na_pass = TRUE,
                   label = "|∆ height| < 100cm",
                   actions = al_strict) %>%
  col_vals_between(vars(shts_diff), -5, 5, na_pass = TRUE,
                   label = "|∆ shoot number| < 5",
                   actions = al_strict) %>%
  interrogate()
```

# Seedlings

## This test is for seedlings that are unusually large.

```{r agent3}
# check seedlings
ha %>%
  filter(recorded_sdlg) %>% 
  create_agent(
    actions = al_default,
    label = "Check seedlings"
  ) %>% 
  col_vals_lt(vars(shts), 3, na_pass = TRUE,
              label = "shoots < 3") %>% 
  col_vals_lt(vars(ht), 30, na_pass = TRUE,
              label = "height < 30cm") %>% 
  interrogate()
```



```{r agent3}
# check seedlings
ha %>%
  filter(recorded_sdlg) %>% 
  create_agent(
    actions = al_default,
    label = "Check seedlings"
  ) %>% 
  col_vals_lt(vars(shts), 3, na_pass = TRUE,
              label = "shoots < 3") %>% 
  col_vals_lt(vars(ht), 30, na_pass = TRUE,
              label = "height < 30cm") %>% 
  interrogate()
```



